Acceleration of Derivative Calculations with Application to Radial Basis Function ­ Finite-Differences on the Intel MIC Architecture

### 2014-06-10 slides/2014_ics/slides-ics.tex
progress on slides

\begin{frame}
  \frametitle{What is Intel MIC}
  Intel Many integrated Core (MIC) is Intel's response to GPUs
  becoming popular in High Performance Computing.

  \begin{block}{What GPUs do well?}
    \begin{itemize}
      \item Get a lot of GFlops by using hundreds of cores
      \item Each core has large SIMD-like abilities
      \item Hide memory latency by using fast cycle context switch
    \end{itemize}
  \end{block}

  \begin{block}{What GPUs do not do well?}
    \begin{itemize}
      \item Alien to program
      \item Poor support for legacy applications
      \item Inter thread communications
      \item Branching
    \end{itemize}
  \end{block}

  Goal of \MIC : do all of it well!
\end{frame}

\begin{frame}
  \frametitle{Overall Architecture}
  \begin{columns}
    \column{18em}    
    \includegraphics[width=18em]{slides-figures/MIC-overview.pdf}
    \column{17em}

    \begin{itemize}
      \item 8 memory controllers
        \begin{itemize}
          \item GDDR5
          \item 2 channels (32-bit)
          \item 5.5GT/s
          \item 352GB/s aggregated peak
            \begin{itemize}
              \item twice the GPU's
              \item but you typically get half
            \end{itemize}
        \end{itemize}
      \item 50+ cores
        \begin{itemize}
          \item 32KB of L1 cache
          \item 512KB of L2 cache
          \item LRU
          \item 8-way associative
        \end{itemize}
      \item 1 PCI-e controller
        \begin{itemize}
          \item to the host (2GB/s guaranteed to memory)
        \end{itemize}
    \end{itemize}
  \end{columns}

@@

\begin{frame}
  \frametitle{Core architecture}
  \begin{columns}
    \column{20em}
    \includegraphics[width=20em,height=20em]{slides-figures/MIC-core.png}%from intelsdg  
    \column{12em}
    \begin{itemize}
      \small
    \item Clocked at 1Ghz
    \item 64-bit
    \item 4 hardware threads
      \begin{itemize}
        \footnotesize
        \item no context switching
        \item no 2 instructions for the same thread
      \end{itemize}
    \item A vectorial unit
      \begin{itemize}
        \footnotesize
        \item (more details on the next slides)
      \end{itemize}
    \item Two instruction pipes:
      \begin{itemize}
        \footnotesize
        \item 2 ALU ops
        \item ALU + MEM ops
        \item ALU + VFP ops
        \item VFP + MEM ops
      \end{itemize}
    \item In-order execution
    \end{itemize}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Vectorial Unit (SIMD) : The classics}

  \begin{columns}
    \column{15em}

    \includegraphics[width=15em]{slides-figures/vect-basic.png}%from intelsdg

    \column{15em}

    {\small 32 512-bits registers (16x32, 8x64)}

    \begin{block}{Binary operators (1 cycle throughput)}
      \small mul, add, sub, and, xor, or, fma
    \end{block}

    \begin{block}{Short Vector Math Library}
      sqrt, rsqrt, log, exp, div
    \end{block}

@@

\begin{frame}
  \frametitle{Getting actual bandwidth}

@@

@@

  \begin{columns}
    \column{.33\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_best-crop.pdf}

@@

    \column{.33\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_worst-crop.pdf}
    \begin{center}
      Worst case\\~
    \end{center}

    \column{.33\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_no_cache-crop.pdf}
    \begin{center}
      Worst case \\no cacheline
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{In terms of flops?}

@@

\section{Implementation}

@@

    \includegraphics[width=15em]{slides-figures/vect-swizzling.png}%from intelsdg

@@

\begin{frame}
  \frametitle{Vectorial Unit (SIMD) : Unaligned loads/stores}

@@

    \begin{block}{Gather}
      Allows to load data from memory at arbitrary offsets. Takes one
      cycle per cache line accessed.
    \end{block}

    \includegraphics[width=\linewidth]{slides-figures/gather.pdf}

    \begin{block}{Scatter}
      Same principle as gather but does store operations.
    \end{block}

@@

    \includegraphics[width=\linewidth]{slides-figures/gather.pdf}

    \begin{block}{Pack}
      Same principle as unpack but does store operations.
    \end{block}    
  \end{columns}  
\end{frame}

@@

  \begin{block}{Observations}
    \begin{itemize}
    \item We aim at over 200Gflop/s in single precision. (\~{}  10\% of peak.)
    \item One instruction in ten must be a Fused Multiply Add.
    \end{itemize}
  \end{block}

  \begin{figure}
    \centering
    \scalebox{.85}{
      \subfigure[Source code]{\includegraphics[clip=true,trim=4.5cm 11cm 4cm 4.2cm,width=.6\linewidth]{slides-figures/ICS-figures/code.pdf}}%
      \subfigure[Content of the vector registers.]{\hspace{-0.07\linewidth}\includegraphics[width=.55\linewidth]{slides-figures/ICS-figures/code-annote.pdf}}
    }
  \end{figure}  
\end{frame}

>>>>

% \begin{frame}
%   \frametitle{What is Intel MIC}
%   Intel Many integrated Core (MIC) is Intel's response to GPUs
%   becoming popular in High Performance Computing.

%   \begin{block}{What GPUs do well?}
%     \begin{itemize}
%       \item Get a lot of GFlops by using hundreds of cores
%       \item Each core has large SIMD-like abilities
%       \item Hide memory latency by using fast cycle context switch
%     \end{itemize}
%   \end{block}

%   \begin{block}{What GPUs do not do well?}
%     \begin{itemize}
%       \item Alien to program
%       \item Poor support for legacy applications
%       \item Inter thread communications
%       \item Branching
%     \end{itemize}
%   \end{block}

%   Goal of \MIC : do all of it well!
% \end{frame}

\begin{frame}
  \frametitle{Intel MIC: Overall Architecture}
  \begin{columns}
    \column{18em}    
    \includegraphics[width=18em]{slides-figures/MIC-overview.pdf}
    \column{17em}

    \begin{itemize}
      \item 8 memory controllers
        \begin{itemize}
          \item GDDR5
          \item 2 channels (32-bit)
          \item 5.5GT/s
          \item 352GB/s aggregated peak
            \begin{itemize}
              \item about same as a K40 
              \item but you typically get half
            \end{itemize}
        \end{itemize}
      \item 50+ cores
        \begin{itemize}
          \item 32KB of L1 cache
          \item 512KB of L2 cache
          \item LRU
          \item 8-way associative
        \end{itemize}
      \item 1 PCI-e controller
        \begin{itemize}
          \item to the host (2GB/s guaranteed to memory)
        \end{itemize}
    \end{itemize}
  \end{columns}

@@

\begin{frame}
  \frametitle{Core architecture}
  \begin{columns}
    \column{20em}
    \includegraphics[width=20em,height=20em]{slides-figures/MIC-core.png}%from intelsdg  
    \column{12em}
    \begin{itemize}
      \small
    \item Clocked at 1Ghz
    \item 64-bit
    \item 4 hardware threads
      \begin{itemize}
        \footnotesize
        \item no context switching
        \item no 2 instructions for the same thread
      \end{itemize}
    \item A vectorial unit
      \begin{itemize}
        \footnotesize
        \item 512-bit registers
        \item 16 SP floats
        \item 8 DP floats
        \item support FMA
      \end{itemize}
    \item Two instruction pipes:
      \begin{itemize}
        \footnotesize
        \item 2 ALU ops
        \item ALU + MEM ops
        \item ALU + VFP ops
        \item VFP + MEM ops
      \end{itemize}
    \item In-order execution
    \end{itemize}
  \end{columns}
\end{frame}

% \begin{frame}
%   \frametitle{Vectorial Unit (SIMD) : The classics}

%   \begin{columns}
%     \column{15em}

%     \includegraphics[width=15em]{slides-figures/vect-basic.png}%from intelsdg

%     \column{15em}

%     {\small 32 512-bits registers (16x32, 8x64)}

%     \begin{block}{Binary operators (1 cycle throughput)}
%       \small mul, add, sub, and, xor, or, fma
%     \end{block}

%     \begin{block}{Short Vector Math Library}
%       sqrt, rsqrt, log, exp, div
%     \end{block}

%   \end{columns}
% \end{frame}

@@

\begin{frame}
  \frametitle{Have to use vector operations to get high bandwidth}

@@

  \tiny Source: PPAM 2013
\end{frame}

@@

  \begin{columns}
    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_best-crop.pdf}

@@

    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_worst-crop.pdf}
    \begin{center}
      Worst case\\~
    \end{center}

    % \column{.33\linewidth}
    % \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_no_cache-crop.pdf}
    % \begin{center}
    %   Worst case \\no cacheline
    % \end{center}
  \end{columns}

  \tiny Also, taking cacheline into account makes a large difference (see paper for details).
\end{frame}

\begin{frame}
  \frametitle{In terms of flops? (Assuming memory bandwidth bound)}

@@

\section{Implementation (4 vectors, 4 matrices)}

\begin{frame}
  \frametitle{Observation on the 4 vectors, 4 matrices case}

  \begin{itemize}
  \item We aim at over 200Gflop/s in single precision. (\~{}  10\% of peak.)
  \item One instruction in ten must be a Fused Multiply Add.
  \item The kernel must be fairly instruction efficient
  \end{itemize}
\end{frame}

@@

    \hspace*{-.1\linewidth}\includegraphics[width=1.2\linewidth]{slides-figures/vect-swizzling.png}%from intelsdg

@@

\begin{frame}
  \frametitle{Vectorial Unit (SIMD) : smarter loads}

@@

    \begin{block}{Gather}
      Allows to load data from memory at arbitrary offsets. Takes one
      cycle per cache line accessed.
    \end{block}
    \includegraphics[width=\linewidth]{slides-figures/gather.pdf}

@@

    \includegraphics[width=\linewidth]{slides-figures/unpack.pdf}

@@

  \begin{figure}
    \centering
    \hspace*{-1em}\scalebox{.95}{
      \subfigure[Source code]{\includegraphics[clip=true,trim=4.5cm 11cm 4cm 4.2cm,width=.6\linewidth]{slides-figures/ICS-figures/code.pdf}}%
      \subfigure[Content of the vector registers.]{\hspace{-0.07\linewidth}\includegraphics[width=.55\linewidth]{slides-figures/ICS-figures/code-annote.pdf}}
    }
  \end{figure}  
\end{frame}
